# Practical Machine Learning Project 

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

## Data 

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.

We read the csv files with na.strings set to appropriate values.

```{r}
ptrain <- read.csv("pml-training.csv", na.strings = c("", "#DIV/0!", "NA"))
ptest <- read.csv("pml-testing.csv", na.strings = c("", "#DIV/0!", "NA"))


# Check dimensions
dim(ptrain)
dim(ptest)
```

## Required Libraries 

```{r, warning=FALSE, message=FALSE}
library(caret)
library(lattice)
library(ggplot2)
library(randomForest)
library(dplyr)
```

## Setting seed for replay

```{r}
set.seed(1234)
```

## Data Preprocessing

We preprocess the data to remove unwanted portion before model creation.

```{r}
# 1. Remove first column
ptrain <- ptrain[-1]

# 2. Identify near zero variance predictors 
nzv_preds <- nearZeroVar(ptrain, saveMetrics = TRUE)

# 3. Remove near zero variance predictors from the dataset
ptrain <- ptrain[, !nzv_preds$nzv]

# 4. Identify predictors with 75% missing values
missing_preds <- sapply(colnames(ptrain), 
                        function(x) if(sum(is.na(ptrain[, x])) > 0.75*nrow(ptrain))
                          return(TRUE)
                        else
                          return(FALSE)
                        )

# 5. Remove predictors with 75% missing values from training dataset
ptrain <- ptrain[, !missing_preds]

dim(ptrain)
```

## Initial Fit of Random Forest Classifier

We use the Random Forest model to fit the training data set. To predict the performance of generated model, we use 10-fold cross-validation.

```{r}
modelFit <- train(classe ~ ., 
                  method = "rf", 
                  data = ptrain, 
                  importance = TRUE, 
                  trControl = trainControl(method = "cv", number = 10)
)
summary(modelFit)
```

## Plot the predictors' contributions to model

```{r, fig.width=8, fig.height=20}
plot(varImp(modelFit))
```

## Details of model

```{r}
modelFit$finalModel
```

We noted that the estimate of error rate of the cross-validated model as 0.04%. The error can be calculated by calculating the ratio of number of wrong classifications to total number of classifications. This ratio is given as (1+3+1+2+1)/(5581+3799+3420+3216+3606) = 0.04% which agrees with the estimate of error.

## Model Prediction

```{r}
# do the prediction on the test data set
predicted <- predict(modelFit, newdata=ptest)

# print the predicted values
print(predicted)
```

## File output
```{r}
# store the predicted values in separate files to be uploaded
# function to write output file
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("./prediction/problem_id_", i, ".txt")
    write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, col.names = FALSE)
  }
}

# call function for each of the predicted values
pml_write_files(predicted)
```